The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging with the Owens Valley Long Wavelength Array

### 2017-06-05 paper.tex
minor progress

However in many cases the study of CMB foregrounds is scientifically interesting in its own right.
Measurements of dust-correlated anomalous microwave emission \citep{1997ApJ...486L..23L} led to the
discovery of spinning dust emission \citep{1998ApJ...508..157D}. Thomson scattering attenuates the
CMB spectrum, but implies a mean redshift to the Epoch of Reionization (EoR) of $8.8^{+1.7}_{-1.4}$
\citep{2016A&A...594A..13P}. Inverse Compton scattering of CMB photons propagating through galaxy
clusters produces an effect known as the Sunyaev-Zel'dovich effect \citep{1970CoASP...2...66S,
1972CoASP...4..173S} that can be used to discover new galaxy clusters \citep{2010ApJ...722.1180V}
and measure the distance to galaxy clusters when combined with X-ray measurements
\citep{2006ApJ...647...25B}.

At redshifts $20 \gtrsim z \gtrsim 7$ the 21 cm hyperfine structure line of neutral hydrogen is
expected to produce a 10 to 100 mK perturbation in the CMB spectrum \citep{2006PhR...433..181F,
2012RPPh...75h6901P}. The amplitude of the perturbation on a given line-of-sight is a function of
the neutral fraction of hydrogen, the baryon overdensity, the spin temperature relative to the CMB
temperature at the given redshift, and the line-of-sight peculiar velocity of the gas.  Current 21
cm cosmology experiments can be broadly separated into two classes: global experiments that are
aiming to detect the spectral signature of the cosmologically redshifted 21 cm transition after
averaging over the entire sky, and power spectrum experiments that incorporate angular information
to attempt to measure the spatial power spectrum of cosmological 21 cm perturbations.

Ongoing global experiments include EDGES, SCI-HI, LEDA, and SARAS 2 \citep{2017arXiv170306647S}.

Ongoing power spectrum experiments include PAPER/HERA \citep{2015ApJ...809...61A,
2016arXiv160607473D}, LOFAR \citep{2017ApJ...838...65P}, and the MWA \citep{2016ApJ...833..102B}.

@@

\todo{discuss more interesting science to be done}

\todo{discuss global experiments too, edges, sci-hi, leda, saras 2}

These low-frequency 21 cm cosmology experiments are limited by the dynamic range they can achieve
against low-redshift sources of emission. The brightness temperature of the galactic synchrotron
emission at high galactic latitudes is measured by \citet{2008AJ....136..641R} as
\begin{equation}
    T \sim 300\,{\rm K} \times \left(\frac{\nu}{150\,{\rm MHz}}\right)^{-2.5}\,.
\end{equation}
Therefore experiments conservatively need to achieve at least 4 orders of dynamic range.  In theory
this kind of dynamic range can be obtained by exploiting the fact that the foreground emission is
predominantly synchrotron emission that has a smooth power law spectrum \todo{citation}.  In
practice the interferometer itself imparts frequency structure into the measurements and therefore
an experiment's ability to separate the foreground signal from the cosmological signal can be
limited by the instrumental spectral smoothness \todo{citation}.

Ideally the foreground removal strategy should be informed by the measured spatial and frequency
structure of all foreground components. This possibility is limited by the availability of suitable
high-fidelity low-frequency sky maps on angular scales ranging from tens of degrees to arcminutes.

@@

Naively one might attempt to solve equation~\ref{eq:basic-imaging} by discretizing, and subsequently
solving the resulting matrix equation. If the interferometer is composed of $N_\text{base}$
baselines, and measures $N_\text{freq}$ frequency channels over $N_\text{time}$ integrations then
the entire data set consists of $N_\text{base}N_\text{freq}N_\text{time}$ complex numbers. If the
sky is discretized into $N_\text{pix}$ pixels then the relevant matrix has dimensions of
$(N_\text{base}N_\text{freq}N_\text{time})\times(N_\text{pix})$. For making single-channel maps with
the Owens Valley LWA this becomes a 5 petabyte array (assuming each matrix element is a 64-bit
complex floating point number).  This matrix equation is therefore prohibitively large, and solving
equation~\ref{eq:basic-imaging} by means of discretization is usually impossible \todo{cite Jayce's
paper (has he published that yet?) where he does this, but explain why his work is the exception
instead of the rule}.

Instead it is common to make mild assumptions that simplify equation~\ref{eq:basic-imaging} and ease
the computational burden in solving for $I_\nu(\hat r)$. For example, when all of the baselines
$\vec b_{ij}$ lie in a plane and the field-of-view is small, equation~\ref{eq:basic-imaging} can be
well-approximated by a two-dimensional Fourier transform.

@@

\citet{2017MNRAS.465.2901Z}

@@

For a transit telescope, the visibilities $V^{ij}_\nu$ are a periodic function of sidereal
time\footnote{
    This is not strictly true. Ionospheric fluctuations and non-sidereal sources (such as the sun)
    will violate this assumption. This paper will, however, demonstrate that the impact on the final
    maps is mild.
}.
Therefore it is a very natural operation to compute the Fourier transform of the visibilities with
respect to sidereal time $\phi\in(0,2\pi]$.
\begin{equation}
    V^{ij}_{m,\nu} = \int_0^{2\pi} V^{ij}_\nu(\phi)\exp\bigg(-im\phi\bigg)\,\d\phi
\end{equation}
The output of this Fourier transform is the set of $m$-modes $V^{ij}_{m,\nu}$ where
$m=0,\,\pm1,\,\pm2,\,\ldots$ is the Fourier conjugate variable to the sidereal time. The $m$-mode
corresponding to $m=0$ is a simple average of the visibilities over sidereal time. Similarly $m=1$
corresponds to the component of the visibilities that varies over half-day timescales. Larger values
of $m$ correspond to components that vary on quicker timescales.

>>>>

However in many cases the study of CMB foregrounds is scientifically interesting in its own right.
Measurements of dust-correlated anomalous microwave emission \citep{1997ApJ...486L..23L} led to the
discovery of spinning dust emission \citep{1998ApJ...508..157D}. Thomson scattering in the IGM
attenuates the CMB spectrum, but the optical depth implies a mean redshift to the Epoch of
Reionization (EoR) of $8.8^{+1.7}_{-1.4}$ \citep{2016A&A...594A..13P}.  The thermal
Sunyaev-Zel'dovich effect \citep{1970CoASP...2...66S, 1972CoASP...4..173S} has been used to discover
new galaxy clusters \citep{2010ApJ...722.1180V} and measure the distance to galaxy clusters when
combined with X-ray measurements \citep{2006ApJ...647...25B}.

\todo{kinetic SZ effect puts an upper limit on the duration of reionization}

At redshifts $20 \gtrsim z \gtrsim 7$ the 21 cm hyperfine structure line of neutral hydrogen is
expected to produce a 10 to 100 mK perturbation in the CMB spectrum \citep{2006PhR...433..181F,
2012RPPh...75h6901P}. The amplitude of this perturbation on a given line-of-sight is a function of
the neutral fraction of hydrogen, the baryon overdensity, the spin temperature relative to the CMB
temperature at the given redshift, and the line-of-sight peculiar velocity of the gas.

@@

Current 21 cm cosmology experiments can be broadly separated into two classes: global experiments
that are aiming to detect the spectral signature of the cosmologically redshifted 21 cm transition
after averaging over the entire sky, and power spectrum experiments that incorporate angular
information to attempt to measure the spatial power spectrum of cosmological 21 cm perturbations.
Ongoing global experiments include EDGES \citep{2010Natur.468..796B, 2017ApJ...835...49M}, LEDA
\todo{cite Price et al. 2017}, BIGHORNS \citep{2015PASA...32....4S}, SCI-HI
\citep{2014ApJ...782L...9V}, and SARAS 2 \citep{2017arXiv170306647S}.  Ongoing power spectrum
experiments include PAPER/HERA \citep{2015ApJ...809...61A, 2016arXiv160607473D}, LOFAR
\citep{2017ApJ...838...65P}, and the MWA \citep{2016ApJ...833..102B}.

Just as for CMB experiments, foreground removal or suppression is an essential component of both
classes of 21 cm cosmology experiments. The brightness temperature of the galactic synchrotron
emission at high galactic latitudes is measured by \citet{2008AJ....136..641R} as
\begin{equation}
    T \sim 300\,{\rm K} \times \left(\frac{\nu}{150\,{\rm MHz}}\right)^{-2.5}\,.
\end{equation}
Therefore experiments conservatively need to achieve at least 4 orders of dynamic range against this
foreground emission before the cosmological signal can be measured. Current foreground removal
methods (for example, \citealt{2012ApJ...756..165P}) rely on the assumption that the foreground
emission (after convolving with the instrumental response) is spectrally smooth. However the
low-frequency radio sky is composed of several components: galactic synchrotron emission, free-free
emission and absorption, supernova remnants, radio galaxies, and a confusing background of point
sources.  Ideally a foreground removal strategy should be informed by the measured spatial and
frequency structure of all foreground components. However, this possibility is limited by the
availability of suitable high-fidelity low-frequency sky maps on angular scales ranging from tens of
degrees to arcminutes.

@@

Naively one might attempt to solve equation~\ref{eq:basic-imaging} by discretizing, and subsequently
solving the resulting matrix equation. If the interferometer is composed of $N_\text{base}$
baselines, and measures $N_\text{freq}$ frequency channels over $N_\text{time}$ integrations then
the entire data set consists of $N_\text{base}N_\text{freq}N_\text{time}$ complex numbers. If the
sky is discretized into $N_\text{pix}$ pixels then the relevant matrix has dimensions of
$(N_\text{base}N_\text{freq}N_\text{time})\times(N_\text{pix})$. For making single-channel maps with
the Owens Valley LWA this becomes a 5 petabyte array (assuming each matrix element is a 64-bit
complex floating point number).  This matrix equation is therefore prohibitively large, and solving
equation~\ref{eq:basic-imaging} by means of discretization is usually impossible although
\citet{2017MNRAS.465.2901Z} demonstrate this technique with the MITEOR telescope.

Instead it is common to make mild assumptions that simplify equation~\ref{eq:basic-imaging} and ease
the computational burden in solving for $I_\nu(\hat r)$. For example, when all of the baselines
$\vec b_{ij}$ lie in a plane and the field-of-view is small, equation~\ref{eq:basic-imaging} can be
well-approximated by a two-dimensional Fourier transform \citep{2001isra.book.....T}.

@@

@@

For a transit telescope, the visibilities $V^{ij}_\nu$ are a periodic function of sidereal
time\footnote{
    This is not strictly true. Ionospheric fluctuations and non-sidereal sources (such as the sun)
    will violate this assumption. This paper will, however, demonstrate that the impact on the final
    maps is mild.
}.
Therefore it is a very natural operation to compute the Fourier transform of the visibilities with
respect to sidereal time $\phi\in[0,2\pi)$.
\begin{equation}
    V^{ij}_{m,\nu} = \int_0^{2\pi} V^{ij}_\nu(\phi)\exp\bigg(-im\phi\bigg)\,\d\phi
\end{equation}
The output of this Fourier transform is the set of $m$-modes $V^{ij}_{m,\nu}$ where
$m=0,\,\pm1,\,\pm2,\,\ldots$ is the Fourier conjugate variable to the sidereal time. The $m$-mode
corresponding to $m=0$ is a simple average of the visibilities over sidereal time. Similarly $m=1$
corresponds to the component of the visibilities that varies over half-day timescales. Larger values
of $m$ correspond to components that vary on quicker timescales.
